Supplementary Materials for CBEA: Competitive Balances for Taxonomic Enrichment Analysis

### 2022-01-27 cbea_revised.tex

Unsurprisingly, our approaches for type I 
However, the trade-off for good type I error control is demonstrably lower power, as shown in Fig~\ref{fig:2}B. In situations where there is no inter-taxa correlation, CBEA still outperforms the wilcoxon rank sum test, however adjusted versions of CBEA did not perform as well as un-adjusted ones. However, in higher correlation scenarios, the difference in power is much more dramatic. At the highest effect size (fold change of 3) and correlation ($\rho = 0.5$), adjusted CBEA was only performing at 50\% power, while unadjusted CBEA and wilcoxon rank sum test were able to reach 80\%. These results indicate that both sparsity and inter-taxa correlation impacts power, with correlation having a much more dramatic impact especially for adjusted versions of CBEA. Most importantly, CBEA demonstrate higher power in all scenarios where type I error is properly controlled.

To further assess the utility of CBEA in classifying samples with enriched sets, we generated AUC scores for different CBEA scores using true labels of whether a sample has an inflated set. This analysis, therefore, assessed the relative ranking of samples using CBEA scores whereby high scores should correspond to samples that are known to be inflated. Fig~\ref{fig:3} presents this result. We compared different variants of CBEA against competing methods in the gene set testing space (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}), as well as the $W$ test statistic from the Wilcoxon rank sum test. Across both simulations (Fig~\ref{fig:3}A) and real-data applications (Fig~\ref{fig:3}B), CBEA scores perform marginally better especially in low effect size situations but did not stand out in most other scenarios. In simulation studies, classification performance was good (around AUC of 0.8) even at high correlation settings, only requiring medium effect sizes (fold change of 2). Notably, the W-statistic provided the least information for classifying samples with inflated taxa.

\subsubsection*{Real data evaluations}  
These observations were replicated when assessed on the semi-labeled gingival data set from the Human Microbiome Project as described in \nameref{methods}. Here, we tested the enrichment of aerobic microbes for each sample using approaches similar to our parametric simulations. As expected in Fig~\ref{fig:2}C, the proportion of falsely rejected hypotheses was high in the naive Wilcoxon test and unadjusted CBEA methods. Conversely, adjusted CBEA controls for false positives adequately at the correct $\alpha$ level of 0.05. Power analysis (Fig~\ref{fig:2}D) showed similar patterns, where unadjusted CBEA methods and the Wilcoxon test have a higher proportion of null hypotheses correctly rejected, however, these results are not useful to a practitioner as the number of falsely rejected hypotheses are also equally high.

\begin{figure}[!h]
    \centering
    \includegraphics[width = \textwidth]{figures/sim_data_ss_auc.png}
    \caption{Classification performance via AUC of CBEA, ssGSEA, GSVA, and Wilcoxon $U$ statistic on simulated data \textbf{(A)} the gingival data set from the Human Microbiome Project  \textbf{(B)} as detailed in \nameref{methods}. Performance scores measure whether scores can highly rank samples that are known to have inflated abundance. In the gingival data set presented in panel \textbf{(B)}, samples from the supragingival site are assumed to have an inflated abundance of aerobic microbes. Error bars are the 95\% DeLong confidence intervals for AUC \cite{delong1988}} 
    \label{fig:3}
\end{figure}

>>>>

Fig~\ref{fig:3} shows results where we assess the phenotype relevance of CBEA results. For power analyses, we observed that adjusted CBEA approaches demonstrate much lower power compared to the Wilcoxon rank-sum test and unadjusted variants. This is surprising given the fact that in statistical significance analyses, the adjusted CBEA approach provides inflated type I error, especially if the normal distribution assumption was chosen. Here, we did not see any differences in performance between distribution choices.

We also evaluate phenotype relevance by assessing whether enriched sets according to ground truth are preferentially ranked higher using CBEA scores. This aspect is captured through computing AUROC by comparing computed enrichment scores and true labels. Consistent with prior approaches, adjusting for correlation did not improve performance, where the AUROC values are equivalent to using the Wilcoxon Rank sum statistic at around 0.5. Unadjusted methods were much better at ranking true enriched sets, however the mean AUROC values are lower than alternate single sample enrichment methods (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}) even though this difference is not significant due to overlapping confidence intervals.  
Unfortunately, evaluating power and score rankings using real data does not allow for observing performance trends when effect sizes, sparsity, and inter-taxa correlation are varied. In equivalent simulation analyses, we observed the opposite effect, where

### 2022-01-27 saving_texts.tex

>>>>

\subsection{Power analyses}
However, the trade-off for good type I error control is demonstrably lower power, as shown in Fig~\ref{fig:2}B. In situations where there is no inter-taxa correlation, CBEA still outperforms the wilcoxon rank sum test, however adjusted versions of CBEA did not perform as well as un-adjusted ones. However, in higher correlation scenarios, the difference in power is much more dramatic. At the highest effect size (fold change of 3) and correlation ($\rho = 0.5$), adjusted CBEA was only performing at 50\% power, while unadjusted CBEA and wilcoxon rank sum test were able to reach 80\%. These results indicate that both sparsity and inter-taxa correlation impacts power, with correlation having a much more dramatic impact especially for adjusted versions of CBEA. Most importantly, CBEA demonstrate higher power in all scenarios where type I error is properly controlled.

To further assess the utility of CBEA in classifying samples with enriched sets, we generated AUC scores for different CBEA scores using true labels of whether a sample has an inflated set. This analysis, therefore, assessed the relative ranking of samples using CBEA scores whereby high scores should correspond to samples that are known to be inflated. Fig~\ref{fig:3} presents this result. We compared different variants of CBEA against competing methods in the gene set testing space (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}), as well as the $W$ test statistic from the Wilcoxon rank sum test. Across both simulations (Fig~\ref{fig:3}A) and real-data applications (Fig~\ref{fig:3}B), CBEA scores perform marginally better especially in low effect size situations but did not stand out in most other scenarios. In simulation studies, classification performance was good (around AUC of 0.8) even at high correlation settings, only requiring medium effect sizes (fold change of 2). Notably, the W-statistic provided the least information for classifying samples with inflated taxa.

\subsubsection*{Real data evaluations}  
These observations were replicated when assessed on the semi-labeled gingival data set from the Human Microbiome Project as described in \nameref{methods}. Here, we tested the enrichment of aerobic microbes for each sample using approaches similar to our parametric simulations. As expected in Fig~\ref{fig:2}C, the proportion of falsely rejected hypotheses was high in the naive Wilcoxon test and unadjusted CBEA methods. Conversely, adjusted CBEA controls for false positives adequately at the correct $\alpha$ level of 0.05. Power analysis (Fig~\ref{fig:2}D) showed similar patterns, where unadjusted CBEA methods and the Wilcoxon test have a higher proportion of null hypotheses correctly rejected, however, these results are not useful to a practitioner as the number of falsely rejected hypotheses are also equally high.
